---
title: "Getting started with RCamelsPE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RCamelsPE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)

camels_path <- system.file(
  "extdata",
  "sample_camels_pe",
  package = "RCamelsPE"
)
```

# Introduction

`RCamelsPE` is an R package designed to read, process, and visualize data from the CAMELS-PE dataset: **Catchment Attributes and Meteorology for Large-sample Studies - Peru**.

This vignette provides a practical introduction to the package using a small example dataset included in the package.

------------------------------------------------------------------------

## Installation

Install the package from CRAN when available:

```{r, eval = FALSE}
install.packages("RCamelsPE")
```

The development version can be installed from GitHub:

```{r, eval = FALSE}
# install.packages("remotes")
remotes::install_github("hllauca/RCamelsPE")
```

Load the package:

```{r setup}
library(RCamelsPE)
```

------------------------------------------------------------------------

## Download the CAMELS-PE dataset

The CAMELS-PE dataset is distributed separately from the package and is available from Zenodo:

<https://doi.org/10.5281/zenodo.20058779>

***The dataset is currently temporarily restricted while the associated scientific article is under preparation.***

The dataset can be downloaded automatically with:

```{r, eval = FALSE}
download_camels_pe(path = tempdir())
```

This function downloads and extracts the dataset, and automatically\
defines the CAMELS-PE dataset path for the current R session.

------------------------------------------------------------------------

## Set the CAMELS-PE dataset path manually

If the dataset has already been downloaded manually, define the local CAMELS-PE path with:

```{r, eval = FALSE}
set_camels_path("path/to/CAMELS-PE")
```

For reproducibility, this vignette uses the small example dataset included in the package:

```{r}
set_camels_path(camels_path)
```


------------------------------------------------------------------------

## Explore the data dictionary

The data dictionary describes available metadata fields, attributes, geospatial layers, time series variables, units, source files, and data sources.

```{r}
dict <- read_dictionary()
head(dict)
```

Time series variables can be inspected with:

```{r}
read_dictionary(category = "timeseries")
```

Topographic attributes can be inspected with:

```{r}
read_dictionary(category = "topographic")
```

------------------------------------------------------------------------

## Explore available stations

The station metadata table contains the available gauging stations and catchments. Each row represents a catchment identified by a unique `gauge_id`.

```{r}
stations <- read_metadata()
head(stations)
```

Select one basin to work with:

```{r}
gauge_id <- 'PE_212900'
```

------------------------------------------------------------------------

## Read geospatial data

CAMELS-PE includes spatial data for catchments and gauges.

```{r}
catchments <- read_geospatial("catchments")
gauges <- read_geospatial("gauges")
```

## Plot catchment location

```{r}
plot_catchments(
  catchments = catchments,
  gauges = gauges,
  gauge_id = gauge_id
)
```

This helps identify the basin location, geographic setting, and relation to other basins.

------------------------------------------------------------------------

## Explore catchment attributes

CAMELS-PE includes multiple groups of attributes describing each basin.

```{r}
topo <- read_attributes("topographic")
head(topo)
```

```{r}
clim <- read_attributes("climatic")
head(clim)
```

Other available attribute groups include:

```{r}
hyd <- read_attributes(type = "hydrological")
land <- read_attributes(type = "landcover")
geo <- read_attributes(type = "geologic")
soil <- read_attributes(type = "soil")
human <- read_attributes(type = "human_intervention")
```

Extract the attributes for the selected basin:

```{r}
topo_sel <- topo[topo$gauge_id == gauge_id, ]
clim_sel <- clim[clim$gauge_id == gauge_id, ]

topo_sel
clim_sel
```

These variables help describe the hydrological system:

-   topography controls flow pathways and energy gradients
-   climate defines water input and atmospheric demand

------------------------------------------------------------------------

## Read hydroclimatic time series

Daily time series may include precipitation, temperature, potential evapotranspiration, observed streamflow, simulated streamflow, solar radiation, and vapor pressure.

```{r}
ts <- read_timeseries(
  gauge_id = gauge_id,
  vars = c("date", "gauge_id", "prec", "flow_obs")
)

head(ts)
```

### Plot precipitation

```{r}
plot_timeseries(ts, variable = "prec", linewidth = 0.4)
```

### Plot streamflow

```{r}
plot_timeseries(ts, variable = "flow_obs")
```

------------------------------------------------------------------------

## Units of streamflow

In CAMELS-PE, observed and simulated streamflow (`flow_obs` and `flow_sim`) are expressed in **mm/day**.

This unit represents runoff depth over the catchment area and allows comparison between basins of different sizes.

In many applications, streamflow is required in **m3/s**. To convert from mm/day to m3/s, the catchment area must be used:

$$
Q_{m3/s} = \frac{Q_{mm/day} \times A_{km2} \times 10^6}{1000 \times 86400}
$$

where:

-   $Q_{mm/day}$ is streamflow in mm/day
-   $A_{km2}$ is catchment area in km2
-   86400 is the number of seconds per day

```{r}
area_km2 <- topo_sel$area

ts$flow_m3s <- ts$flow_obs * area_km2 * 1e6 / (1000 * 86400)

plot_timeseries(ts, variable = "flow_m3s")
```

------------------------------------------------------------------------

## Compare multiple catchments

```{r}
ts_multi <- read_timeseries(
  gauge_id = stations$gauge_id[1:2],
  vars = c("date", "gauge_id", "prec")
)
```

```{r}
plot_timeseries(ts_multi, variable = "prec", facet = TRUE)
```

------------------------------------------------------------------------

## Filter a specific catchment

```{r}
plot_timeseries(
  ts_multi,
  variable = "prec",
  gauge_id = stations$gauge_id[1]
)
```

------------------------------------------------------------------------

# Typical workflow

```{r, eval = FALSE}
library(RCamelsPE)

# Optional: download dataset from Zenodo
# download_camels_pe(path = tempdir())

set_camels_path("path/to/CAMELS-PE")

read_dictionary(category = "timeseries")

stations <- read_metadata()

gauge_id <- stations$gauge_id[1]

ts <- read_timeseries(
  gauge_id = gauge_id,
  vars = c("date", "gauge_id", "prec", "flow_obs")
)

plot_timeseries(ts, variable = "flow_obs")
```

# Summary

With `RCamelsPE`, users can:

-   access hydroclimatic data
-   explore basin characteristics
-   visualize time series and spatial data
-   compare multiple catchments

This makes the package suitable for hydrological analysis, teaching, model development, and large-sample hydrology studies.

# Notes

To run the examples with real data, download the CAMELS-PE dataset using `download_camels_pe()` or manually from Zenodo.

The `download_camels_pe()` function automatically defines the dataset path for the current R session.

If the dataset was downloaded manually, or in a previous R session, define the local dataset path using `set_camels_path()`.
