---
title: "Getting started with RCamelsPE"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with RCamelsPE}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Introduction

`RCamelsPE` is an R package designed to read, process, and visualize data from the CAMELS-PE dataset:\
**Catchment Attributes and Meteorology for Large-sample Studies – Peru**.

This vignette provides a practical introduction to the package, guiding users through a typical workflow:

-   explore available basins
-   inspect their physical and climatic characteristics
-   analyze hydroclimatic time series

------------------------------------------------------------------------

## Installation

Before using the package, you need to install it from GitHub. This step only needs to be done once.

```{r}
# install.packages("remotes")
remotes::install_github("hllauca/RCamelsPE")
```

Once installed, load the package into your R session:

```{r setup}
library(RCamelsPE)
```

------------------------------------------------------------------------

## Set the CAMELS-PE dataset path

CAMELS-PE is not included in the package, so you must specify the location of the dataset on your local system. This allows all functions to automatically locate the data.

```{r}
set_camels_path("C:/Users/hllauca/Documents/CAMELS-PE")
```

------------------------------------------------------------------------

## Explore available stations

We begin by loading the metadata, which contains all available gauging stations. This table is the entry point to the dataset.

```{r}
stations <- read_metadata()

head(stations)
```

Each row represents a catchment, identified by a unique `gauge_id`.

We select one basin to work with:

```{r}
gauge_id <- 'PE_472935F2'

gauge_id
```

------------------------------------------------------------------------

## Read geospatial data

CAMELS-PE also includes spatial data for catchments and gauges.

These layers allow geographic visualization.

```{r}
catchments <- read_geospatial("catchments")
gauges     <- read_geospatial("gauges")
```

## Plot catchment location

Finally, we visualize the selected basin in its spatial context.

```{r}
plot_catchments(
  catchments = catchments,
  gauges = gauges,
  gauge_id = gauge_id
)
```

This helps understand:

-   where the basin is located

-   its geographic setting

-   its relation to other basins

------------------------------------------------------------------------

## Explore catchment attributes

CAMELS-PE includes multiple groups of attributes describing each basin. These attributes are essential for understanding hydrological behavior.

```{r}
topo <- read_attributes("topographic")
head(topo)
```

```{r}
clim <- read_attributes("climatic")
head(clim)
```

Other available attribute groups include:

```{r}
hyd   <- read_attributes(type = "hydrological")
land  <- read_attributes(type = "landcover")
geo   <- read_attributes(type = "geologic")
soil  <- read_attributes(type = "soil")
human <- read_attributes(type = "human_intervention")
```

Now we extract the attributes corresponding to the selected basin:

```{r}
topo_sel <- topo[topo$gauge_id == gauge_id, ]
clim_sel <- clim[clim$gauge_id == gauge_id, ]

topo_sel
clim_sel
```

These variables provide insight into the system:

-   **Topography** controls flow pathways and energy gradients

-   **Climate** defines water input and atmospheric demand

------------------------------------------------------------------------

## Read hydroclimatic time series

Next, we load daily time series for the selected catchment.

This includes key hydroclimatic variables such as precipitation, temperature, evapotranspiration, and streamflow. Additional variables are also available in CAMELS-PE (e.g., air temperature, potential evapotranspiraton, simulated streamflow, solar radiation and vapor pressure), and can be selected using the `vars` argument.

```{r}
ts <- read_timeseries(
  gauge_id = gauge_id,
  vars = c("date", "gauge_id", "prec", "flow_obs")
)

head(ts)
```

### Plot precipitation

We first explore precipitation, which is the main driver of the hydrological system.

```{r}
plot_timeseries(ts, variable = "prec", linewidth = 0.4)
```

This plot helps identify:

-   seasonal patterns

-   variability across years

-   extreme events

### Plot streamflow

Now we analyze observed streamflow, which integrates all basin processes.

```{r}
plot_timeseries(ts, variable = "flow_obs")
```

Streamflow reflects:

-   rainfall input

-   soil moisture dynamics

-   groundwater contributions

-   river routing

## Units of streamflow

In CAMELS-PE, both observed and simulated streamflow (`flow_obs` and `flow_sim`) are expressed in **mm/day**.

This unit represents runoff depth over the catchment area and is commonly used in hydrology because it allows comparison between basins of different sizes.

However, in many applications, streamflow is required in **m³/s**.\
To convert from mm/day to m³/s, the catchment area must be used:

$$
Q_{m^3/s} = \frac{Q_{mm/day} \times A_{km^2} \times 10^6}{1000 \times 86400}
$$

Where:

-   $Q_{mm/day}$ = streamflow in mm/day\
-   $A_{km^2}$ = catchment area in km²\
-   86400 = seconds per day

```{r}
# Read catchment area from topographic attributes
area_km2 <- topo_sel$area

# Convert flow from mm/day to m3/s
ts$flow_m3s <- ts$flow_obs * area_km2 * 1e6 / (1000 * 86400)

plot_timeseries(ts, variable = "flow_m3s")
```

------------------------------------------------------------------------

## Compare multiple catchments

To understand spatial variability, we can analyze several basins at once.

```{r}
ts_multi <- read_timeseries(
  gauge_id = c("PE_472A9204", "PE_210003"),
  vars = c("date", "gauge_id", "prec")
)
```

Plot them using faceting:

```{r}
plot_timeseries(ts_multi, variable = "prec", facet = TRUE)
```

Each panel represents a different catchment, allowing comparison of hydrological regimes.

------------------------------------------------------------------------

## Filter a specific catchment

Even when working with multiple basins, you can easily focus on one.

```{r}
plot_timeseries(
  ts_multi,
  variable = "prec",
  gauge_id = "PE_472A9204"
)
```

------------------------------------------------------------------------

# Typical workflow

A minimal workflow with `RCamelsPE` is:

```{r}
library(RCamelsPE)

set_camels_path("C:/Users/hllauca/Documents/CAMELS-PE")

stations <- read_metadata()

gauge_id <- stations$gauge_id[1]

ts <- read_timeseries(
  gauge_id = gauge_id,
  vars = c("date", "gauge_id", "prec", "flow_obs")
)

plot_timeseries(ts, variable = "flow_obs")
```

# Summary

With `RCamelsPE`, users can:

-   access hydroclimatic data quickly

-   explore basin characteristics

-   visualize time series and spatial data

-   compare multiple catchments

This makes the package suitable for:

-   hydrological analysis

-   teaching and training

-   model development

-   large-sample hydrology studies

# Notes

This vignette uses `eval = FALSE` because CAMELS-PE must be available locally.\
To run the examples, download the dataset and set the correct path using `set_camels_path()`.
